Understanding the Impact of Mortgage State and Income on Household Poverty Status

### 2024-04-11 paper/paper.qmd
some changes

---
title: "Understanding the Impact of Mortgage State and Income on Household Poverty Status"
subtitle: "An Analysis of Household Poverty Status in the United States in 2019"
author: 
  - Mary Cheng
thanks: "Code and data are available at: https://github.com/marycx/us_poverty_analysis_2019.git"
date: today
date-format: long
abstract: "This paper examines household poverty status in the United States in 2019 using the 2019 Revised Supplemental Poverty Measure (SPM) Research Dataset from the United States Census Bureau. Using logistic regression analysis, we investigate the influence of household income level after tax and mortgage state on poverty status. Our findings reveal a higher likelihood of poverty for households' total annual earning less than $50000 and for renters compared to property owners. All households with incomes exceeding $100000 are not in poverty. These insights emphasize the need for targeted interventions to address socioeconomic inequalities and promote financial well-being among US households."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

\newpage
# Introduction
Over the last few decades, addressing poverty has remained a persistent issue in the United States, leading to the implementation of many policies in an effort to mitigate this problem [@Kearney_Harris]. In 2019, the United States experienced the culmination of an economic expansion that followed the Great Recession. The robustness of the U.S. economy during this period was evidenced by low poverty rates, indicating a notable decrease in economic hardship. The poverty rate in the United States was estimated to be 10.5% in 2019, down from 11.8% in 2018, marking the fifth consecutive year of decline [@Semega_2022]. In order to reduce poverty further, it is important for us to analyze why some US household are in poverty in 2019 and how various demographic and financial factors may has an influence on it. We analyze the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset from the United States Census Bureau (USCB) [@Bureau_2022]. USCB is an agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy [@uscb]. Supplemental Poverty Measure was developed to address numerous concerns of official poverty measure critics, and its intent is to provide an improved statistical picture of poverty [@Gesumaria_2015]. We build a prediction model for households poverty status using 2019 Revised SPM dataset. This allows us to discover trends and address potential financial and societal situations that may cause the poverty status for some US households in 2019.

In this paper, a logistic regression model is used to predict the poverty status of households in the United States in 2019, with data from the 2019 Revised Supplemental Poverty Measure Research dataset. Logistic regression is a great choice since it is used to predict binary outcomes, such as poverty status (in poverty or not in poverty). Our analysis focuses on estimating the likelihood of household being in poverty, based on various demographic and financial factors captured in the SPM dataset. We select 17 data features for analysis. After some data cleaning, we summarize 17 features into 2 variables: 1) mortgage state and 2) household annual total income after tax, plus benefits and minus expenses. The estimand in this paper is the number of households who are in poverty in reality. However, it is difficult to measure the exact number of households who are in poverty since there are millions of people in the United States and not all of them will be assessed due to various difficulties. Therefore, in this paper, we attempt to estimate the estimand using a logistic regression model which is trained using sample dataset from the 2019 Revised Supplemental Poverty Measure Research Data from USCS.

@@

The intercept ($\alpha$) and coefficients ($\beta_1$, $\beta_2$) are assigned informative prior distributions to regularize the model. Specifically, we assume a normal distribution with a mean of 0 and a standard deviation of 2.5 for each parameter.

We chose this modeling approach for several reasons. Firstly, logistic regression is ideal for binary outcome variables, making it appropriate for analyzing poverty status. Also, Bayesian methods enable us to integrate prior knowledge and uncertainty into the analysis, resulting in more reliable estimates of the model parameters.

While alternative modeling approaches like linear regression were considered, we selected Bayesian logistic regression due to the binary nature of our outcome variable, poverty status.

@@

>>>>

---
title: "Understanding the Impact of Mortgage State and Income on Household Poverty Status"
subtitle: "An Analysis of Household Poverty Status in the United States in 2019"
author: 
  - Mary Cheng
thanks: "Code and data are available at: https://github.com/marycx/us_poverty_analysis_2019.git"
date: today
date-format: long
abstract: "This paper examines household poverty status in the United States in 2019 using the 2019 Revised Supplemental Poverty Measure Research Dataset from the United States Census Bureau. Using logistic regression analysis, we investigate the influence of household income level and mortgage state on poverty status. Our findings reveal a higher likelihood of poverty for households' total annual earning less than $50000 and for renters compared to property owners. All households with incomes exceeding $100000 are not in poverty. This implies that household mortgage status and income plays an important role in poverty status. These insights emphasize the need for targeted interventions to address socioeconomic inequalities and promote financial well-being among US households."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

\newpage
# Introduction
Over the last few decades, the United States have implemented many policies, in an effort to address and mitigate poverty [@Kearney_Harris]. In 2019, the United States experienced the culmination of an economic expansion that followed the Great Recession. The robustness of the U.S. economy during this period was evidenced by low poverty rates, which was estimated to be 10.5% in 2019, down from 11.8% in 2018, marking the fifth consecutive year of decline [@Semega_2022]. In order to reduce poverty further, it is important for us to analyze why some US household are in poverty in 2019. We analyze the 2019 Revised Supplemental Poverty Measure (SPM) Research dataset from the United States Census Bureau (USCB) [@Bureau_2022]. USCB is an agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy [@uscb]. Supplemental Poverty Measure was developed to address numerous concerns of official poverty measure critics, providing an improved statistical picture of poverty [@Gesumaria_2015]. We build a prediction model for households poverty status using 2019 Revised SPM dataset. We primarily look at two factors: mortgage status and income in each household. This allows us to look at the question whether potential financial factors may cause the poverty status for some US households in 2019.

In this paper, a logistic regression model is used to predict the poverty status of households in the United States in 2019, with data from the 2019 Revised Supplemental Poverty Measure Research dataset. Logistic regression is a great choice since it is used to predict binary outcomes, such as poverty status (in poverty or not in poverty). Our analysis focuses on estimating the likelihood of household being in poverty, based on various financial factors captured in the SPM dataset. We select 17 data features for analysis. After some data cleaning, we summarize 17 features into 2 variables: 1) mortgage state and 2) household annual total income after tax, plus government benefits and minus expenses. The estimand in this paper is the number of households who are in poverty in reality. However, it is difficult to measure the exact number of households who are in poverty since there are millions of people in the United States and not all of them will be assessed due to various difficulties. For example, due to regional issues, some people who live in rural areas are not included in the survey. Therefore, in this paper, we attempt to estimate the estimand using a logistic regression model which is trained using sample dataset from the 2019 Revised Supplemental Poverty Measure Research Data from USCS.

@@

The intercept ($\alpha$) and coefficients ($\beta_1$, $\beta_2$) are assigned informative prior distributions to regularize the model. Specifically, we assume a normal distribution with a mean of 0 and a standard deviation of 2.5 for each parameter. Setting the prior to 0 assumes there is no poverty in any household regardless of their mortgage or income. Then, when we factor in real observations, we can see how things shift. If the distribution shifts left (less than 0), then we know that particular mortgage or income state is more likely to result in household poverty. This approach gives us a clear way to see how data affects the financial situation of households

We chose this modeling approach for several reasons. Firstly, logistic regression is ideal for binary outcome variables, making it appropriate for analyzing poverty status. Also, Bayesian methods enable us to integrate prior knowledge and uncertainty into the analysis, resulting in more reliable estimates of the model parameters. While alternative modeling approaches like linear regression were considered, we selected Bayesian logistic regression due to the binary nature of our outcome variable: poverty status.

@@

## Policy Implications
Broaden the discussion to include potential policy implications and recommendations based on your findings. How can policymakers use this information to address household poverty?
